---
title: "Experimental Design and ANOVA"
author: "Alex Sánchez"
output:
   html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    theme: cerulean
    highlight: textmate
    number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width = 12, fig.height = 8, cache = FALSE,
                      echo = TRUE, warning = FALSE, message = FALSE)
```

#  Introduction

This document is a refresher about the relation between Experimental Design and ANOVA, mainly through the linear model.

The idea beghind it is as follows.

- __Experimental Designs__ describe the way to assign treatments to experimental units so that the effect of these treatments can be properly investigated. 

This can be done in many ways that go from the simplest experimental design, CRD, to many other designs.

- The _Completely Randomized Design_ is intended to investigate the effect of two or more treatments without additinal sources of variation.

- The _Randomized Block Design_ is used when, in addition to the treatments there is one or more additional sources of _uncontrolled_ variation in the samples. Researchers can randomlñy assign treatments to experimental units but not blocks -each unit belongs to a given block and this canot be changed- 

- _Factorial Designs_ are used when there are two or more sources of variation. Each source can be independently be assigned to each experimental unit, and interaction between them can be studied if replicates are available.

- _Repeated Measures Designs_ appear  when in the previous situations one take multiple measurements on each experimental unit.
    
Each experimental design can be __described__ using a linear model and

The data obtained from a study where a given experimental has been used can be __analyzed__ using its associated Analysis of the Variance model.

# Examples

## Completely Randomized design


```{r}
RESP<-c(96,99,100,104,84,91,90,75,80,90,70,90,84,76,78,78,87,67,66,76)
TREAT<-rep(1:4, each=5)
dades<-data.frame (TREAT, RESP)
dades$TREAT <- as.factor(TREAT)
kableExtra::kable(dades)
```


```{r}
attach (dades)
write.csv2(dades, file="dadesCRD.csv2", row.names=TRUE)


model <- RESP ~ TREAT

aov1<-aov (model, data=dades)
summary(aov1)

model.tables(aov1)
model.tables(aov1,type="means") # Mitjanes dels grups

(hsd=TukeyHSD(aov1,which="TREAT",conf=0.95))
plot(hsd)

### Check ANOVA model assumptions

opt<- par(mfrow=c(2,2))
plot(aov1)
par(opt)

### Clean memory
detach(dades)
rm(dades)
```


## Randomized block design

```{r}
regeneration<-c(34.98,41.22,36.94,39.97,40.89,46.69, 46.65, 41.90, 42.07, 49.42, 52.68, 42.91, 37.18, 45.85, 40.23, 39.20, 37.99, 41.99, 37.61, 40.45, 34.89, 50.15, 44.57, 43.29)
dades<-expand.grid(medium=1:4, Tx =1:6)
dades<-cbind(dades,regeneration)
dades$medium<-as.factor(dades$medium)
dades$Tx<-as.factor(dades$Tx)
write.csv2(dades, "dadesRBD.csv2", row.names=TRUE)
kableExtra::kable(dades)
```


```{r}
attach(dades)
model.1<- regeneration ~ medium + Tx
aov.1<-aov (model.1, data=dades)
summary(aov.1)

### If blocking had been ignored

model.0<- regeneration ~  Tx
aov.0<-aov (model.0, data=dades)
summary(aov.0)

### Check ANOVA model assumptions

opt<- par(mfrow=c(2,2))
plot(aov1)
par(opt)

### Clean memory
detach(dades)
rm(dades)
```


## Factorial design

```{r}
dades<-expand.grid(replica=1:5, tratamiento = c("Control", "Trat"),cepa=c("CD1", "CBA"))
WBC<-c( 3, 1.7, 1.5,2, 2.1, 1.9, 1.9, 3.5, 1.2, 2.3, 1.9, 2.6, 1.4, 1.6, 1.1, 0.4, 0.2, 0.1, 0.4, 0.3)
dades$cepa <-as.factor(dades$cepa)
dades$tratamiento <-as.factor(dades$tratamiento)
dades <- cbind(dades, WBC)
write.csv2(dades, "dadesFACTD.csv2", row.names=F)
kableExtra::kable(dades)
```


```{r}
attach(dades)

model.1<- WBC ~ tratamiento + cepa # this model ignores interaction
aov.1<-aov (model.1, data=dades)
summary(aov.1)

model.2<- WBC ~ tratamiento + cepa + cepa:tratamiento
aov.2<-aov (model.2, data=dades)
summary(aov.2) # this shows that interaction had to be considered

interaction.plot(cepa, tratamiento, WBC)

### Check ANOVA model assumptions
opt<- par(mfrow=c(2,2)); plot(aov1); par(opt)

### Clean memory
detach(dades); rm(dades)
```


one single factor with two or more levels